The imprint of global magma oceans on exoplanet demographics

### 2023-03-24 src/tex/ms.tex
discuss multi-planet systems

\section{Baseline Model}\label{sec:met_baseline}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
%Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
%To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically. %in simulated exoplanet observations.
Our basic methodology was as follows:
We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
\todo[inline]{\bioverse\ will need a short introduction before the modifications are mentioned. Probably should get a paragraph in the intro}
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}
In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.

@@

\subsection{Detectability of the runaway greenhouse transition}
\todo[inline]{Mention multiplanet systems.}
We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
However, even assuming capable instrumentation and an optimal survey design, geophysical and demographic unknowns may jeopardize this undertaking.
The detectability of the runaway greenhouse threshold is thus a function of both survey design and astrophysical circumstances.

@@

\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}

@@

\subsection{Impact of assumptions on our findings}
\todo[inline]{Caveats\\
context: \citet{Turbet2019} Sect. 3, \citet{Bower2019}.\\
''present bracketing cases; or to discuss why the discountinuity in the results would not be affected by changes that led to continuous offsets in the results. Or both.''\\
- not yet explored: how sensitive are we on the assumed planet population? $\exists$ danger of false positive detection on occurrence rate ``ridges'' like the Neptune desert? Currently we have a relatively smooth occurrence rate density in $R_P - S$ space based on the power-law fits in \citet{Bergsten2022}. We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
The prospects for detecting the runaway greenhouse transition depends on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
We can only approximate all this here, and we therefore seek to assess the sensitivity of our results on the assumptions we have made.

>>>>

\section{Baseline Model}\label{sec:met_baseline}
%The purpose of this paper is to demonstrate the feasibility of measuring the demographic feature introduced above with near-future exoplanet missions.
%Central to our procedure is to explore the detectability of population-level trends with future exoplanet surveys, and here the trend in question is the proposed radius discontinuity.
%To enable a quantitative assessment of this detectability, we injected the signal into a simulated planet sample before observing it with simulated surveys.
The goal of this study is to determine -- for different configurations of near-future exoplanet surveys -- the confidence level with which the runaway greenhouse threshold can be detected statistically. %in simulated exoplanet observations.
Our basic methodology was as follows:
We expanded the \bioverse\ framework~\citep{Bixel2020,Bixel2021}\footnote{\bioverse\ is actively maintained and documented open source software written in Python. Its latest version and documentation can be found at \url{https://github.com/danielapai/bioverse}.} to generate synthetic samples of stars that host planets according to the observed exoplanet demographics.
We then adapted planetary bulk properties as predicted from models of runaway greenhouse atmospheres, simulated observations of the planets, and computed Bayesian evidences in favor of the runaway greenhouse hypothesis (see diagram in Fig.~\ref{fig:flowchart}).\todo[inline]{make sure rgh hypothesis is mentioned before, or rephrase}
\begin{figure*}
    \begin{centering}
        \includegraphics[width=\hsize]{figures/flowchart.pdf}
        \caption{Workflow of our hypothesis testing with \bioverse. First, we generate a sample of stars and populate them with planets based on \kepler\ demographics.
            A fraction of them are then assigned a runaway greenhouse climate based on the model described in Sect.~\ref{sec:met_rghmodel}.
        We then simulate an exoplanet survey, whereby selection effects and detection biases are introduced. Finally, we test the runaway greenhouse hypothesis based on data from the survey simulation.
        By iterating through these steps, we compute the statistical power of testing the hypothesis for different survey designs.}
        \label{fig:flowchart}
    \end{centering}
\end{figure*}
In this section, we review the source of the stellar sample, the modeled luminosity evolution, the generation of a synthetic planet sample, as well as the orbital parameters of the planets.
An overview of our key assumptions and model parameters can be found in Table~\ref{tab:params_table}.

@@

\subsection{Detectability of the runaway greenhouse transition}
We showed in Sect.~\ref{sec:res_testability} that under favorable conditions, a sufficiently large (\var{N_optimistic} planets) photometric survey is likely to detect the demographic imprint of the runaway greenhouse transition and accurately constrain the associated threshold instellation.
However, even assuming capable instrumentation and an optimal survey design, geophysical and demographic unknowns may jeopardize this undertaking.
The detectability of the runaway greenhouse threshold is thus a function of both survey design and astrophysical circumstances.

@@

\subsubsection{OPTIONAL...: Atmospheric signatures}\label{sec:dis_atmospheres}
\todo[inline]{TIM: Ideas?\\
discuss potential atmospheric signatures of magma oceans. e.g.: H/O ratios of sub-Neptunes could be low (because H is in the melt) (e.g. Tim's recent talk)}
\todo[inline]{``Can a smaller sample size's diagnostic value be increased if atmospheric compositional information can be added, i.e., PLATO + ARIEL (or similar combo)?''}
\todo[inline]{also: use atmospheric characterization to identify small planet cores with thick H/He envelopes that contribute to diluting the signal (aka background population). These will be amenable for transmission spec due to large scale heights (Moran2018, \citet{Turbet2019})}

\subsubsection{Detection in multi-planet systems}
As suggested by \citet{Turbet2019}, an alternative approach for detecting the runaway greenhouse-induced radius inflation is to search for its ``local'' imprints in multi-planet systems.
Systems harboring planets on both sides of the transition, such as \mbox{TRAPPIST-1}~\citep{Gillon2016a,Gillon2017a,Luger2017c,Agol2021}, may show the predicted abrupt radius and density change, provided the initial volatile content was sufficient and complete desiccation has not yet occurred.
While degeneracies remain in interpreting bulk density fluctuations within individual systems~\citep[e.g.,][]{Turbet2020,Dorn2021}, the detection of a consistent pattern in several such systems could be a convincing statistical evidence of the runaway greenhouse transition.

@@

\subsection{Impact of assumptions on our findings}
\todo[inline]{Caveats\\
context: \citet{Turbet2019} Sect. 3, \citet{Bower2019}.\\
- not yet explored: how sensitive are we on the assumed planet population? $\exists$ danger of false positive detection on occurrence rate ``ridges'' like the Neptune desert? Currently we have a relatively smooth occurrence rate density in $R_P - S$ space based on the power-law fits in \citet{Bergsten2022}. We could, with relatively moderate effort, run a sanity check on an alternative sample (e.g., the more structured Bern model populations, or the population in the Genesis database)}
The prospects for detecting the runaway greenhouse transition depends on astro- and geophysical factors, as well as on the specific instrumentation and survey strategy of a particular mission.
We can only approximate all this here, and we therefore seek to assess the sensitivity of our results on the assumptions we have made.
